---
title: "Modeling spatially resolved omics with mistyR"
author:
- name: Jovan Tanevski
  affiliation: 
  - Heidelberg University and Heidelberg University Hospital, Heidelberg, Germany
  - Jožef Stefan Institute, Ljubljana, Slovenia
  email: jovan.tanevski@uni-heidelberg.de
date: "`r Sys.Date()`"
package: mistyR
output:
  BiocStyle::html_document:
    toc_float: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The use of mistyR is conceptualized around building a workflow for analysis of spatial omics data by four classes of functions:

-   View composition

-   Train models

-   Result processing

-   Plotting

To construct a workflow `r Githubpkg("saezlab/mistyR")` is designed with the use of pipe operators (for example the operator `%>%` from `r CRANpkg("magrittr")`) for chaining functions in mind.

When loading `r Githubpkg("saezlab/mistyR")` please consider configuring a `future::multisession()` parallel execution plan. `r Githubpkg("saezlab/mistyR")` will then use all available cores for execution of computationally demanding functions. It is recommended that the user modifies the `future::plan()` according to their needs.

```{r setup, message=FALSE}
library(mistyR)
library(magrittr)
library(future)

plan(multisession)
```

```{r}

```

# View composition

# Train models

# Results processing

# Plotting

---

# See also {-}

## More vignettes {-}

`vignette("mistySeurat")`

`vignette("mistySpatialExperiment")`

## Paper {-}

*`r format(citation("mistyR"), "textVersion")`*

# Session info {-}

Here is the output of `sessionInfo()` at the point when this document was compiled:

```{r info, echo=FALSE}
sessionInfo()
```
