---
title: "Modeling spatially resolved omics with MISTy"
author:
- name: Jovan Tanevski
  affiliation: 
  - Heidelberg University & Heidelberg University Hospital, Heidelberg, Germany
  email: jovan.tanevski@uni-heidelberg.de
date: "`r Sys.Date()`"
package: MISTy
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{MISTy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The use of MISTy is conceptualized around building a workflow for analysis using four classes of functions:

-   View composition

-   Train models

-   Result processing

-   Plotting

```{r setup}
library(MISTy)
library(magrittr)
library(future)
```

MISTy is designed with the use of pipe operators (for example the operator `%>%` from `r CRANpkg("magrittr")`) to construct a workflow by chaining functions in mind.

When loading MISTy please consider configuring a `future::multisession()` parallel 
execution plan. MISTy will then use all available cores for execution of 
computationally demanding functions. It is recommended that the user modifies the 
`future::plan()` according to their needs.

# View composition

# Train models

# Results processing

# Plotting

# Session info

Here is the output of `sessionInfo()` at the point when this document was compiled:

```{r info, echo=FALSE}
sessionInfo()
```

